An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-09 paper.tex
Fixes to 2.1

Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.  Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
parameters implied by the data:
\begin{equation}
  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
\end{equation}
where $p(M_i, \vtheta_i |d)$ is the posterior distribution for the
model parameters $\vtheta_i$ implied by the data in the context of
model $M_i$, $p(\vtheta_i|M_i)$ is the prior
probability of the model parameters that 
represents our beliefs before accumulating any of the data $d$, and
$p(d)$ is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
the $\vtheta_i$.  This implies that the evidence is equal to
\bel{evidence}
  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
\end{equation}
where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
\begin{equation}
p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
\end{equation}
where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
\be
p(d)=\sum_i p(d|M_i) p(M_i).
\ee

@@

A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
\be
p(\vtheta^(i+1))=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
\ee
where $p(\vtheta^{(i)} \to \vtheta^{(i+1}))$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.  In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(i)})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(i+1)})=p(\vtheta|d,M)$.

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^(i)$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^(i) \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\be
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1})=\vtheta^{(i)}$.
 \end{enumerate}

>>>>

Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.  Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
parameters implied by the data:
\begin{equation}
  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
\end{equation}
where $p(\vtheta_i |d, M_i)$ is the posterior distribution for the
model parameters $\vtheta_i$ implied by the data in the context of
model $M_i$, $p(\vtheta_i|M_i)$ is the prior
probability of the model parameters that 
represents our beliefs before accumulating any of the data $d$, and
$p(d|M_i)$, called the evidence, is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
the $\vtheta_i$.  This implies that the evidence is equal to
\bel{evidence}
  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
\end{equation}
where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
\begin{equation}
p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
\end{equation}
where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
\be
p(d)=\sum_i p(d|M_i) p(M_i).
\ee

@@

A Markov chain has the property that the probability distribution of the next state can depend only on the current state, not on the past history:
\be
p(\vtheta^{(i+1)})=\int_{V} d\vtheta^{i} p(\vtheta^{(i)} \to \vtheta^{(i+1)}) p(\vtheta^{(i+1)}),
\ee
where $p(\vtheta^{(i)} \to \vtheta^{(i+1)})$ depends only on $\vtheta^{(i)}$ and $\vtheta^{(i+1)}$. 
An additional requirement for an MCMC arises from the fact that the desired distribution is the equilibrium distribution.  In other words, if we assume that state $(i)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(i)})=p(\vtheta|d,M)$ , then the next state $(i+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(i+1)})=p(\vtheta|d,M)$.

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\be
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1})=\vtheta^{(i)}$.
 \end{enumerate}
